\subsection{Compactness}

\NI First-order logic has compactness: a set $S$ of sentences has a
model exactly when every finite subset of $S$ does \cite[Chapter
  4.3]{EndertonHB:matinttl}. What about eremic logic? 

We can prove compactness of modal logics using the standard
translation from modal to first-order logic \cite{BlackburnP:modlog}:
we start from a set of modal formula such that each finite subset has
a model. We translate the modal formulae and models to first-order
logic, getting a set of first-order formulae such that each finite
subset has a first-order model. By compactness of first-order logic we
obtain a first-order model of the translated modal formulae. Then we
translate that first-order model back to modal logic, obtaining a
model for the original modal formulae, as required.

Unfortunately we cannot do the same with the translation from eremic
to first-order logic presented in the previous section. The problem
are the first-order models termed 'junk' above: they do not correspond
to eremic transistion systems.  For example the constraint $s
\TRANS{a} t$ implies $a \in \lambda(s)$ might be violated. After all,
merely having signature $\SSS$ is not strong enough a constraint. The
target language of the translation from the previous section is not
expressive enough to express such constraints.  As we have no reason
to belive that the first-order model whose existence is guarnateed by
compactness isn't 'junk', we cannot use this translation.

We solve this problem with a second translation from eremic to
first-order logic, this time into a more expressive fragment were we
can constrain first-order models enough to ensure that they always can
be translated back to eremic logic.

The second embedding translates eremic logic to two-sorted first-order
logic. The many-sorted first-order signature $\SSS'$ has two sorts,
states and actions. The action constants are given by $\Sigma$. There
are no state constants. The signature has a nullary predicate $\top$,
an binary predicate $\ALLOWED{}{\cdot}{\cdot}$ and a ternary predicate
$\ARROWTWO{}{\cdot}{\cdot}{\cdot}$.  $\SSS'$ has no functions, i.e.~is
a relational signature.  The intended interpretation should should be
clear.

With the target logic in place, we can now present a second encoding
$\SEMBTWO{\phi}_x$ of eremic formulae.

\begin{eqnarray*}
  \SEMBTWO{\top}_x & \ = \ & \top
     \\
  \SEMBTWO{\phi \AND \psi}_x & = & \SEMBTWO{\phi}_x \AND \SEMBTWO{\psi}_x
     \\
  \SEMBTWO{\langle a \rangle \phi}_x & = & \exists^{state} y.(\ARROW{a}{x}{y} \AND \SEMBTWO{\phi}_y)
     \\
  \SEMBTWO{\fBang A}_x & = & \forall^{action} a.(\ALLOWED{}{x}{a} \IMPLIES a \in A) 
\end{eqnarray*}

\NI Here we use $\exists^{state}$ to indicate that the quantifier ranges
of the sort of states, and $\forall^{action}$ for a quantifier ranging
over actions. The expression $a \in A$ is a shorthand for the
first-order formula
\[
   a = a_1 \OR a = a_2 \OR \cdots \OR a = a_n
\]
assuming that $A = \{a_1, ..., a_n\}$.

Note that the translation above could be restricted to a two-variable
fragment. Moreover, the standard reduction from many-sorted to
one-sorted first-order logic, does not increase the number of
variables used (although predicates are added, one per sort)
\cite{EndertonHB:matinttl}. For simplicity we will not consider this
matter further.

Before we can state and prove a correspondence theorem for
$\SEMBTWO{\phi}_x$ along the lines of Theorem
\ref{correspondence:theorem:1}, we must also translate eremic
  transition systems $\SEMBTWO{\LLL}$.

\begin{definition}
Let $\LLL = (S, \rightarrow, \lambda)$ be an eremic transition
system. Clearly $\LLL$ gives rise to an $\SSS'$-model $\SEMBTWO{\LLL}$
as follows.
\begin{itemize}

\item For each constant $a \in \Sigma$, $a^{\SEMBTWO{\LLL}}$ is $a$ itself.

\item The sort of states is interpreted by the set $S$.

\item The sort of actions is interpreted by the set $\Sigma$.

\item The relation symbols are interpreted as follows.

  \begin{itemize}

    \item $\top^{\SEMBTWO{\LLL}}$ always holds.

    \item $\ALLOWED{\SEMBTWO{\LLL}}{s}{a}$ holds whenever $a \in \lambda(s)$.

    \item $\ARROWTWO{\SEMBTWO{\LLL}}{s}{a}{t}$ holds whenver $s \TRANS{a} t$.

  \end{itemize}
\end{itemize}
\end{definition}


\begin{theorem}[correspondence theorem]\label{correspondence:theorem:2}
Let $\phi$ be an eremic formula and $\MMM = (\LLL, s)$ an eremic
model.
\[
   \MMM \models \phi \quad  \text{iff} \quad \SEMBTWO{\LLL} \models_{x \mapsto s} \SEMBTWO{\phi}_x.
\]
\end{theorem}
\begin{proof}
The proof proceeds by induction on the structure of $\phi$ and is
similar to that of Theorem \ref{correspondence:theorem:2}.  \qed
\end{proof}

\NI We now use the translation $\SEMBTWO{\phi}_x$ to show that eremic
logic must also have compactness. The key steps in the proof are
simple, following standard techniques from modal logic
\cite{BlackburnP:modlog}:

\begin{enumerate}

\item Choose a set $T$ of eremic formulae, such that each finite
  subset $T'$ of $T$ has an eremic model $(\LLL, s)$.

\item Using the translations gives a set $\SEMBTWO{T} =
  \{\SEMBTWO{\phi}\ |\ \phi \in T\}$ of first-order formulae such that
  each finite subset has a first-order model $\SEMBTWO{\LLL}$.

\item By compactness of first-order logic, we can find a first-order
  model $\CAL{M}$ of $\SEMBTWO{T}$.

\item\label{compactness:step:4} Convert $\CAL{M}$ into an eremic transition system
  $\CAL{M}^{\sharp}$ such that $(\CAL{M}^{\sharp}, s) \models T$.

\end{enumerate}

\NI The problematic step is (\ref{compactness:step:4}), for how would
we know that the model $\CAL{M}$ can be converted back to an eremic
transition system? Why should $\CAL{M}$ exhibit admissibility or
well-sizedness?  The mere fact that $\CAL{M}$ is a first-order model
of signature $\SSS'$ is not strong enough to guarantee these
properties.  We deal with this in two ways. To ensure admissibility,
we define a formula that guarantees that models satisfying the formula
are admissible.
\begin{eqnarray*}
   \phi_{admis} 
      & \ =\ &
   \forall^{state} s.\forall^{action} a.\forall^{state} t.( \ARROWTWO{}{s}{a}{t} \IMPLIES \ALLOWED{}{s}{a})
\end{eqnarray*}

\begin{lemma}\label{compactness:lemma:23399}
If $\LLL$ is an eremic transistion system then $\SEMBTWO{\LLL} \models
\phi_{admis}$.
\end{lemma}

\begin{proof}
Straightforward from the definitions.
\qed
\end{proof}

\NI We can now add, without changing satisfiability, $\phi_{admis}$ to
any set of first-order formulae that has a model that is the
translation of an eremic model.

We deal with the absence of well-sizedness by 

First we add a formula $\phi_a$ for each action $a \in \Sigma$.
\begin{eqnarray*}
   \phi_{a} 
      & = &
   \exists^{action} c. a = c 
\end{eqnarray*}

\begin{lemma}\label{compactness:lemma:666}
If $\LLL$ is an eremic transistion system then $\SEMBTWO{\LLL} \models
\phi_{a}$ for all $a \in \Sigma$.
\end{lemma}

\begin{proof}
Straightforward from the definitions.
\qed
\end{proof}

\begin{definition}
Let $\LLL = (S, \rightarrow, \lambda)$ be an eremic transition system
and $X$ a set, condidered to contain actions. The \emph{restriction of
  $\LLL$ to $X$}, written $\LLL \setminus X$ is the eremic model $(S,
\rightarrow', \lambda')$ where $\rightarrow' = \{(s, a, t) \in
\rightarrow \ |\ a \notin X\}$, and for all states $s$ we set:
\[
   \lambda'(s) 
        =
   \begin{cases}
       \lambda(s) \setminus  X & \text{whenever}\ \lambda(s) \neq \Sigma \\
       \Sigma & \text{otherwise}
   \end{cases}
\]

\end{definition}

\begin{lemma}\label{compactness:lemma:1717}
Let $\phi$ be an eremic formula and $X$ be a set such that no action
occuring in $\phi$ is in $X$. Then:
\[
   (\LLL, s) \models \phi
      \quad\text{iff}\quad
   (\LLL \setminus X, s) \models \phi.
\]
\end{lemma}
\begin{proof}
By straightforward induction on the structure of $\phi$.
\qed
\end{proof}

\begin{definition}
Let $\CAL{M}$ be a first-order model for the signature $\SSS'$.
We construct an eremic transition system
$\CAL{M}^{\sharp} = (S, \rightarrow, \lambda)$.
\begin{itemize}

\item The actions $\Sigma$ are given by the $\CAL{M}$ interpretation of actions.

\item The states $S$ are given by the $\CAL{M}$ interpretation of states.

\item The reduction relation $s \TRANS{a} t$ holds exactly when
  $\ARROWTWO{\CAL{M}}{s}{a}{t}$.

\item The function $\lambda$ is given by the following clause:
  $\lambda(s) = A$ exactly when $\mathsf{Restrict}_{A}^{\CAL{M}}(x)$
  holds.

\end{itemize}

\end{definition}

\begin{theorem}[correspondence theorem]\label{correspondence:theorem:223}
Let $\CAL{M}$ be a first-order model for the signature $\SSS'$ such that
\begin{itemize}

\item $\CAL{M} \models \phi_{admis}$.
 
\item $\CAL{M} \models \phi_{a}$ for all $a \in \Sigma$.

\end{itemize}

\NI Then $\CAL{M}^{\sharp}$ is an eremic transition system with
actions $\Sigma$. Moreover, for all eremic formulae $\phi$ with
actions from $\Sigma$ we have:
\[
   \CAL{M} \models_{x \mapsto s} \SEMBTWO{\phi}_x 
        \quad  \text{iff} \quad 
   (\CAL{M}^{\sharp} \setminus X, s) \models \phi.
\]
\end{theorem}
Here $X$ is the set of all elements in the universe of $\CAL{M}$ interpreting
actions that are not in $\Sigma$.
\begin{proof}
The proof proceeds by induction on the structure of $\phi$. \textbf{To do.}
\end{proof}

\begin{definition}
Let $T$ be a set of eremic formulae, and $\MMM$ an eremic model.  We
write $\MMM \models T$ provided $\MMM \models \phi$ for all $\phi \in
T$.  We say $T$ is \emph{satisfiable} provided $\MMM \models T$.
\end{definition}

\begin{theorem}[Compactness of eremic logic]
A set $T$ of eremic formulae is satisfiable iff each finite subset of
$T$ is satisfiable.
\end{theorem}
\begin{proof}
For the non-trivial direction, let $T$ be a set of eremic formulae
such that any finite subset has an eremic model. Define 
\[
  \SEMBTWO{T} 
     \ =\ 
  \{\SEMBTWO{\phi}\ |\ \phi \in T\} 
     \qquad\qquad
  T^*
     \ =\ 
  \SEMBTWO{T} \cup \{\phi_{admis}\} \cup \{ \phi_a\ |\ a \in \Sigma\}
\]
which both are sets of first-order formulae. Clearly each finite subset $T'$ of 
$T^*$ has a first-order model. Why? First consider the subset $T'_{EL}$ of $T'$
which is given as follows.
\[
   T'_{EL} \ =\ \{ \phi \in T\ |\ \SEMBTWO{\phi} \in T' \}
\]
Since $T'_{EL}$ is finite, by assumption there is an eremic model 
\[
   (\LLL, s) \models T'_{EL}
\]
which means we can apply Theorem \ref{correspondence:theorem:223} to get
\[
   \SEMBTWO{\LLL} \models_{x \mapsto s} \SEMBTWO{T'_{EL}},
\]
By construction $T' \setminus \SEMBTWO{T'_{EL}} \subseteq
\{\phi_{admis}\} \cup \{ \phi_a\ |\ a \in \Sigma\}$, so all we have to
show for $T'$ to have a model is that
\[
    \SEMBTWO{\LLL} \models_{x \mapsto s} \{\phi_{admis}\} \cup \{ \phi_a\ |\ a \in \Sigma\},
\]
but that is a direct consequence of Lemmas
\ref{compactness:lemma:23399} and \ref{compactness:lemma:666}.  That
means each finite subset of $T^*$ has a model and by appealing to
compactness of first-order many-sorted logic (which is an immediate
consequence of compactenss of one-sorted first-order logic
\cite{EndertonHB:matinttl}), we know there must be a first-order model
$\CAL{M}$ of $T^*$, i.e.
\[
   \CAL{M} \models T^*.
\]
Since $\CAL{M} \models \phi_{admis}$ and $\CAL{M} \models \phi_a$ for all actions $a$, 
we can apply Theorem \ref{correspondence:theorem:223} that also
\[
   (\CAL{M}^{\sharp} \setminus X, s) \models T
\]
where $X$ is the set of all actions in $\CAL{M}^{\sharp}$ that are not
in $\Sigma$. Hence $T$ is satisfiable.
 \qed
\end{proof}


