\section{\ELFULL}\label{coreEL}

In this section we introduce the syntax and semantics of \ELFULL{}
(hereafter \ELABR{}).  

\subsection{Syntax}
\label{elsyntax}
\NI Syntactically, eremic logic is a multi-modal logic with one new
operator.

\begin{definition} Let $\Sigma$ be a non-empty set of \emph{actions}.
Actions are ranged over by $a, a', a_1, b, ...$, and $A$ ranges over
finite subsets of $\Sigma$. The \emph{formulae}, ranged over by $\phi,
\psi, \xi ...$, of \ELABR{} are given by the
following grammar.

\begin{GRAMMAR}
  \phi 
     &\quad ::= \quad & 
  \TRUE 
     \VERTICAL 
  \phi \AND \psi
     \VERTICAL 
  \MAY{a}{\phi}
     \VERTICAL 
  \fBang A 
\end{GRAMMAR}
\end{definition}

\NI Before presenting models and the satisfaction relation, we sketch
the meaning of formulae informally. $\TRUE$ is truth and $\phi \AND
\psi$ is the conjunction of formulae $\phi$ and $\psi$. Clearly
$\MAY{a}{\phi}$ is the may-modality which asserts that the current
state may do action $a$, and in doing $a$ transitions evolves into a
now state at which $\phi$ holds. Tantum $A$, from Latin ``tantum''
meaning ``only'' and written $!A$, is the key novelty of eremic logic.
$!A$ means that if $!A$ holds at the current state, then the only
modalities $\MAY{a}{}$ available at the corrent state are those with
$a \in A$.

The set $\Sigma$ of actions must be non-empty to avoid that the logic
is trivial. The restriction to finite subsets of action substantially
simplifies the meta-theory of the logic.

We assume that $\MAY{a}{\phi}$ binds more tightly than conjunction, so
$\MAY{a}{\phi} \AND \psi$ is short for $(\MAY{a}{\phi}) \AND \psi$.
We often abbreviate $\MAY{a}{\TRUE}$ to $\MAY{a}{}$. We define falsity
$\FALSE$ as $!\emptyset \AND \MAY{a}{}$ where $a$ is an arbitrary
action. Note that in the absence of conventional negation, we cannot
readily define disjunction, implication, or must-modalities by de
Morgan duality. 

\begin{convention}
From now on we assume a fixed set $\Sigma$ of actions, except where
stated otherwise.
\end{convention}

\subsection{Semantics}

\NI The semantics of eremic logic is close to Hennessy-Milner logic
\cite{HennessyM:alglawfndac} but uses \emph{deterministic} transition systems, as
described in Section \ref{preliminaries}, augmented with labels on
states.

%% \begin{definition}

%% \NI We say $\LLL$ is \emph{deterministic} if $x \TRANS{a} y$ and $x \TRANS{a} z$ imply that $y = z$. Otherwise $\LLL$ is
%% \emph{non-deterministic}.
%% \end{definition}

\begin{definition}
An \emph{eremic transition system} is a triple $\LLL = (S,
\rightarrow, \lambda)$, where $(S, \rightarrow)$ is a deterministic
labelled transition system over $\Sigma$, and $\lambda$ is a function
from states to sets of actions (not necessarily finite), subject to
the following constraints:
\begin{itemize}

\item For all states $s \in S$ it is the case that $ \{a \fOr \exists
  t \; s \xrightarrow{a} t\} \subseteq \lambda(s)$. We call this
  condition \emph{admissibility}.

\item For all states $s \in S$, $\lambda (s)$ is either finite or
  $\Sigma$. We call this condition \emph{well-sizedness}.

\end{itemize}
\end{definition}

\NI The intended interpretation is that $\lambda(w)$ is the set of
allowed transition symbols emanating from $w$.  The $\lambda$ function
is the semantic counterpart of the $!$ operator.  The admissibility
restriction is in place because transitions $s \TRANS{a} t$ where $a
\notin \lambda(s)$ would be saying that an $a$ action is possible at
$s$ but at the same time prohibit $a$ actions at that state.
Well-sizedness is not a fundamental restriction but rather a
convenient trick. Essentially eremic transition systems have two kinds
of nodes:

\begin{itemize}

\item Nodes $s$ without restrictions on outgoing transitions. Those are
  labelled with $\lambda ( s) = \Sigma$.

\item Nodes $s$ with restriction on outgoing transitions. Those are
  labelled by a finite set $\lambda ( s)$ of actions.

\end{itemize}

\NI Defining $\lambda$ on all nodes and not just on those with
restrictions makes some definitions and proofs slightly easier.

As with other modal logics, satisfaction of formulae is defined
relative to a state in the ambient eremic transition system, giving
rise to the following definition.

\begin{definition}
A \emph{eremic model}, ranged over by $\MMM, \MMM', ...$, is a pair
$(\LLL, s)$, where $\LLL$ is an eremic transition system $(S,
\rightarrow, \lambda)$, and $s$ is a state from $S$. We call $s$ the
\emph{root} of the model.

An eremic model with root $r$ is a \emph{tree}, provided the
underlying transition system is a tree with root $r$.
\end{definition}

\NI With eremic models at hand, we can finally formalise the
satisfaction relation for eremic formulae.

\begin{definition}\label{ELsatisfaction}
The \emph{satisfaction relation} $\MMM \models \phi$ is defined
inductively by the following clauses, where we assume that $\MMM =
(\LLL, s)$ and $\LLL = (S, \rightarrow, \lambda)$.
\[
\begin{array}{lclcl}
  \MMM & \models & \top   \\
  \MMM & \models & \phi \AND \psi &\ \mbox{ iff } \ & \MMM  \models \phi \mbox { and } \MMM \models \psi  \\
  \MMM & \models & \langle a \rangle \phi & \mbox{ iff } & \text{there is transition } s \xrightarrow{a} t \mbox { such that } (\LLL, t) \models \phi  \\
  \MMM & \models & \fBang A &\mbox{ iff } & \lambda(s) \subseteq A
\end{array}
\]
\end{definition}

\NI The first three clauses are standard. The last clause enforces the
intended meaning of $!A$: the available modalities in the model are
\emph{at least as constrained} as required by $!A$. They may even be
more constrained if the inclusion $\lambda(s) \subseteq A$ is
proper. We note in the case where the set $\Sigma$ of actions is
infinite, allowing $\lambda(s)$ to return arbitrary inifinite sets in
addition to $\sigma$ does not make a difference because $A$ is finite
by construction, so $\lambda(s) \subseteq A$ can never hold when
$\lambda(s)$ is infinite. 

\input{figure:elSmall}

We continue with concrete examples.  The model in Figure
\ref{figure:elSmall} satisfies all the following formulae, amongst
others:
\[
\begin{array}{lclclclcl}
\MAY{a} &\qquad&
\MAY{a} \MAY{b} &\qquad&
\MAY{a} \fBang \{b,c\} &\qquad&
\MAY{a} \fBang \{b,c,d\} &\qquad&
\MAY{c} \\[1mm]
\MAY{c} \fBang \{\} &&
\MAY{c} \fBang \{a\} &&
\MAY{c} \fBang \{a,b\} &&
\MAY{a} \land \MAY{c} &&
\MAY{a} (\MAY{b} \land \fBang \{b,c\}
\end{array}
\]

\NI Here we assume, as we do with all subsequent such figures, that
the top note is the root.  The same model does \emph{not} satisfy any
of the following formulae
\[
\MAY{b} \qquad
\fBang \{a\} \qquad
\fBang \{a, c\} \qquad
\MAY{a} \fBang \{b\} \qquad
\MAY{a} \MAY{c} \qquad
\MAY{a} \MAY{b} \fBang \{c\} 
\]

\NI Figure \ref{threemodels} shows various models of $\MAY{a} \MAY{b}$
and Figure \ref{more models} shows one model that does, and one that
does not, satisfy the formula $\fBang \{a,b\}$.  Both models validate
$!\{a, b, c\}$.

\ELABR{} does not have the operators $\neg, \lor, $ or $\Rightarrow$.
This has the following two signification consequences.  First, every
satisfiable formula has a unique (up to isomorphism) simplest model.
In Figure \ref{threemodels}, the left model is the unique simplest
model satisfying$\MAY{a} \MAY{b}$.  We will clarify below that model
simplicity is closely related to the process theoretic concept of
similarity, and use the existence of unique simplest models to
construct a \emph{linear-time} decision procedure.

\input{figure:elAndBang:models}

\input{figure:elAndBang:moreMdels}

Secondly, EL is different from other logics in that there is an
asymmetry between tautologies and contradictories: logics with
conventional negation have an infinite number of non-trivial
tautologies as well as an infinite number of contradictories.  In
contrast, because eremic logic has no negation or disjunction
operator, it is expressively limited in the tautologies it can
express: top and conjunctions of top are its sole tautologies. On the
other hand, the tantum operator, enables an infinite number of
contradictories to be expressed.  For example:
\[
   \MAY{a} \land \fBang \{\} \qquad
   \MAY{a} \land \fBang \{b\} \qquad
   \MAY{a} \land \fBang \{b, c\} \qquad
   \MAY{b} \land \fBang \{\} \qquad
\]


%% \begin{definition} Let $\Gamma$ be an arbitrary set of formulae. We
%% say \emph{$\Gamma$ semantically implies $\phi$}, written $\Gamma
%% \models \phi$, provided for all eremic models $\MMM$ if it is the
%% case that $\MMM \models \Gamma$ implies $\MMM \models \phi$.
%% \richard{We do not define $\models \Gamma$ for a set $\Gamma$, only
%% for individual formulae - do we want to change this definition to
%% be of the form $\phi \models \psi$?}  \end{definition}

\NI The semantic consequence relation is standard.\martin{actually
  it's not because the premise can have only one formula. Maybe we
  should elaborate upon this? I think it's intimately related to the
  absence of negation. In classical logic we can use $\Gamma \models
  \phi$ because only a finite bit already implies $ \phi$ and that can
  easily be proven. In EL ... not so much}

\begin{definition} 
 We say the formula $\phi$ \emph{semantically implies} $\psi$, written $\phi
   \models \psi$, provided for all eremic models $\MMM$ if it is the
   case that $\MMM \models \phi$ then also $\MMM \models \phi$.
\end{definition}

\begin{example}
\ELABR{} shares with other (multi)-modal logics the following
implications:
\begin{eqnarray*}
\MAY{a} \MAY{b} \models \MAY{a} 
 \qquad\qquad
\MAY{a} (\MAY{b} \land \MAY{c}) \models \MAY{a} \MAY{b}
\end{eqnarray*}
As \ELABR{} is restricted to deterministic models, it also
validates the following formula:
\begin{eqnarray*}
\MAY{a} \MAY{b} \land \MAY{a} \MAY{b}  \models \MAY{a} (\MAY{b} \land \MAY{c})
\end{eqnarray*}
\ELABR{} also validates all implications in which the set of constraints is relaxed from left to right. For example:
\begin{eqnarray*}
\fBang \{\} \models \fBang \{a\} 
 \qquad\qquad
\fBang \{\} \models \fBang \{a, b\} 
\end{eqnarray*}
\end{example}
