

\subsubsection{Incompatibility as a Fundamental Semantic Relation}
According to Brandom's, the notion of incompatibility is more than just a pre-logical version of negation. 
It also has a \emph{modal} component:
\begin{quote}
The notion of incompatibility can be thought of as a sort of conceptual vector-product of a \emph{negative} component and a \emph{modal} component. It is \emph{non}-com\emph{possibility}\footnote{\cite{brandom} p.126.}.
\end{quote}
Incompatibility is the fundamental master concept that Brandom uses to make sense of the notion of \emph{object} and \emph{agent}.

First, \emph{object-hood}.
Treating ``$A$ is $\phi$'' and ``$B$ is $\psi$'' as incompatible (where $\phi$ and $\psi$ are incompatible predicates) \emph{just is} what it is to treat $A$ and $B$ as the same object:
\begin{quote}
It is not impossible for two different objects to have incompatible properties - say, being copper and electrically insulating. What is impossible is for \emph{one and the same object} to do so. 
Objects play the conceptual functional role of \emph{units of account for alethic modal incompatibilities}. 
A single object just is what cannot have incompatible properties (at the same time).
\end{quote}
It is only because we are continually resolving incompatibilities that we are able to represent an external (mind-independent) object at all:
\begin{quote}
Shouldering the responsibility of repair and rectification of incompatible commitments is what one has to \emph{do} in order to be taking one's claims to be \emph{about} an objective world
\end{quote}
Second, \emph{agent-hood}.
Just as an object \emph{cannot} have incompatible properties, just so a subject \emph{should not} ascribe incompatible properties to something.
To say that two sentences are incompatible is to say that an agent who holds one \emph{should not} hold the other.
\begin{quote}
Objects play the conceptual functional role of \emph{units of account for alethic modal incompatibilities}. 
It is an essential individuating feature of the metaphysical categorical sortal metaconcept \emph{object} that objects have the metaproperty of \emph{modally} repelling incompatibilities...
And, in a parallel fashion, subjects too are individuated by the way they normatively `repel' incompatible commitments.
It is not impermissible for two different subjects to have incompatible commitments - say, for me to take the coin to be copper and you take it to be an electrical insulator. What is impermissible is for \emph{one and the same subject} to do so. Subjects play the conceptual functional role of \emph{units of account for deontic normative incompatibilities}. 
That is, it is an essential individuating feature of the metaphysical categorical sortal metaconcept \emph{subject} that subjects have the metaproperty of \emph{normatively} repelling incompatibilities. A single subject just is what ought not to have incompatible commitments (at the same time)\footnote{\cite{brandom} p.192.}.
\end{quote}

\NI Sellars and Brandom follow Hegel's \emph{Wissenschaft der Logic}
\cite{HegelGWF:wisdlog} in seeing material incompatibility as a
foundational concept from which even the supposedly primitive ideas of
object and agent can be explicated.  If incompatibility is to fill
this fundamental load-bearing role, we should explore all possible
ways of formalising it.  The tantum operator is the simplest
formalisation we could find in which incompatible propositions can be
expressed.





\subsection{A ``Simple'' Logic}
\Cathoristic{} is a simple logic of atomic sentences.  It contains no
``complex'' logical connectives.  
By ``complex'', we mean logical
connectives that, like disjunction, generate sentences that can be
satisfied in many different distinct ways.  
Negation, disjunction,
implication, existential quantification all create formulae that can
be satisfied in different ways.  
 Disjunction
fails to be simple because there are two equally simple models of
$\phi \lor \psi$.  Similarly, negation fails to be simple because
there are two equally simple models of $\neg (\phi \land \psi)$.
Conjunction is the \emph{only} connective of propositional logic that
is simple, according to this criterion.

Let us define a logic as ``simple'' if the set of satisfying
models of every sentence has a unique least upper bound. 
We saw in Theorem \ref{theorem:decision} of Section \ref{boundedlattice} that \cathoristic{} is simple, according to this definition.
It is because of this simplicity that it has a quadratic-time decision procedure.
\martin{I think this is still only tangentally related to the rest of the conclusion}
\richard{What do you want to do? Remove it? Put it somewhere else?}
 We close this paper with a discussion of related work and open
problems.

