\section{Introduction}\label{introduction}

Natural language is full of incompatible alternatives.
If Pierre is the current king of France, then nobody else can simultaneously fill that role.
A traffic light can be green, amber or red - but it cannot be more than one colour at a time.
Mutual exclusion is a natural and ubiquitous concept.
\FOL{} can represent mutually exclusive alternatives, of course.
To say that Pierre is the only king of France, we can write, following Russell:
\[
king(france, pierre) \land \forall x . (king(france, x) \rightarrow x = pierre).
\]
To say that a particular traffic light, $tl$, is red - and red is its only colour - we could write:
\[
colour(tl, red) \land \forall x . colour(tl, x) \rightarrow x = red.
\]
In this approach, incompatibility is a \emph{derived} concept, reduced to 
a combination of universal quantification and identity.  
\FOL{}, in other words, uses relatively complex machinery to express a
simple concept:
\begin{itemize}

\item Quantification's complexity comes from the
  rules governing the distinction between free
  and bound variables.
  
\item Identity's complexity comes from the infinite collection of axioms required to formalise the
  indiscernibility of identicals.

\end{itemize}

\NI The costs of quantification and identity, such as a larger proof
search space, have to be borne every time one expresses a sentence that excludes others - even
though incompatibility does not, prima facie, appear to have anything to do
with the free/bound variable distinction, or require the full power of 
the identity relation.

This paper introduces an alternative approach, where
exclusion is expressed directly, as a first-class concept.
 \Cathoristic{}\footnote{``Cathoristic'' comes from the Greek
  $\kappa \alpha \theta o \rho \acute{\i} \zeta \epsilon i \nu$: to impose narrow
  boundaries. We are grateful to Tim Whitmarsh for suggesting this
  word.} is the simplest logic we could find in which incompatible
statements can be expressed.  
It is a multi-modal logic, a variant of Hennessy-Milner logic,
that replaces negation with a new logical primitive
\[
   !A
\]
pronounced \emph{tantum}\footnote{``Tantum'' is Latin for ``only''.}
$A$. Here $A$ is a finite set of alternatives, and $!A$ says that the
alternatives in $A$ exhaust all possibilities.  For example:
\begin{eqnarray*}
\fBang \{green, amber, red\}
\end{eqnarray*}
states that nothing but $green$, $amber$ or $red$ is possible.  Our
logic uses modalities to state facts, for example $\MAY{amber}{}$
expresses that $amber$ is currently the case.  The power of the logic
comes from the conjunction of modalities and tantum. For example
\[
   \MAY{amber}{}\ \AND\ !\{green, amber, red\} 
\]
expresses that $amber$ is currently the case and $red$ as well as
$green$ are the only two possible alternatives to $amber$.  Any
statement that exceeds what tantum $A$ allows, like
\[
   \MAY{blue} \ \AND\ !\{green, amber, red\},
\]
is necessarily false.  When the only options are green, amber, or red,
then blue is not permissible.  Now to say that Pierre is the only king
of France, we write:
\[
\MAY{king}\MAY{france}(\MAY{pierre} \land \fBang \{pierre\}).
\]
Crucially, \cathoristic{}'s representation involves no
universal quantifier and no identity relation.  It is a purely
propositional formulation.  To say that the traffic
light is currently red, and red is its only colour, we write:
\[
\MAY{tl} \MAY{colour} (\MAY{red} \land !\{red\}).
\]
This is simpler, both in terms of representation length and
computational complexity, than the formulation in \fol{} given on the
previous page.
%% \[
%% colour(tl, red) \land \forall x . colour(tl, x) \rightarrow x = red
%% \]
Properties changing over time can be expressed by adding extra
modalities that can be understood as time-stamps.  To say that that
the traffic light was red at time $t_1$ and amber at time $t_2$, we
can write:
\[
   \MAY{tl} \MAY{colour} (\MAY{t_1} (\MAY{red} \land !\{red\}) \land \MAY{t_2} (\MAY{amber} \land !\{amber\}))
\]
Change over time can be expressed in first-order logic with bounded
quantification - but modalities are succinct and avoid introducing
bound variables.

Having claimed that incompatibility is a natural logical concept, not
easily expressed in first-order logic\footnote{We will precisify this
  claim in later sections; (1) first-order logic's representation of
  incompatibility is longer in terms of formula length than
  \cathoristic{}'s (see Section \ref{incompatiblepredicatesinfol});
  and (2) logic programs in \cathoristic{} can be optimised to run
  significantly faster than their equivalent in \fol{} (see Section
  \ref{optimizingpreconditions}).}, we will now argue the following:

\begin{itemize}

\item Incompatibility is conceptually prior to negation.

\item Negation arises as the weakest form of incompatibility.

\end{itemize}

\subsection{Material incompatibility and negation}

\NI Every English speaker knows that
\begin{quote}
``Jack is alive'' is incompatible with ``Jack is dead''
\end{quote}

\NI But \emph{why} are these sentences incompatible? The orthodox
position is that these sentences are incompatible because of the
following general law:
\begin{quote}
If someone is alive, then it is not the case that they are dead
\end{quote}
Recast in first-order logic:
\[
\forall x. ( alive(x) \IMPLIES \neg dead(x) ).
\]

\NI In other words, according to the orthodox position, the
incompatibility between the two particular sentences depends on a
general law involving universal quantification, implication and
negation.

Brandom \cite{brandom2} follows Sellars in proposing an alternative explanation: ``Jack
is alive'' is incompatible with ``Jack is dead'' because ``is alive''
and ``is dead'' are \emph{materially incompatible} predicates.  They
claim we can understand incompatible predicates even if we do
not understand universal quantification or negation.  
Material incompatibility is conceptually prior to logical negation.

Imagine, to make this vivid, a primitive people speaking a primordial
language of atomic sentences\footnote{In this paper, we define a sentence as \emph{atomic} if it does
  not contain another sentence as a syntactic constituent.}. These people can express sentences
that \emph{are} incompatible.  But they cannot express \emph{that}
they are incompatible.  They recognise when atomic sentences are
incompatible, and see that one sentence entails another - but their
behaviour outreaches their ability to articulate it.

Over time, these people \emph{may} advance to a more sophisticated
language where incompatibilities are made explicit, using a negation
operator - but this is a later (and optional) development:
\begin{quote}
[If negation is added to the language], it lets one say that two
claims are materially incompatible:``If a monochromatic patch is red,
then it is not blue.'' That is, negation lets one make explicit in the
form of claims - something that can be said and (so) thought - a
relation that otherwise remained implicit in what one practically did,
namely treat two claims as materially
incompatible\footnote{\cite{brandom} pp.47-48}.
\end{quote}

\NI But before making this optional
explicating step, our primitive people understand incompatibility
without understanding negation.  If this picture of our primordial
language is coherent, then material incompatibility is conceptually
independent of logical negation.

Now imagine a modification of our primitive linguistic practice in
which no sentences are ever treated as incompatible.  If one person
says ``Jack is alive'' and another says ``Jack is dead'', nobody
counts these claims as \emph{conflicting}.  The native speakers never
disagree, back down, retract their claims, or justify them. They just
say things.  Without an understanding of incompatibility, and the
variety of behaviour that it engenders, we submit (following Brandom)
that there is insufficient richness in the linguistic practice for
their sounds to count as assertions.  Without material
incompatibility, their sounds are just \emph{barks}.
\begin{quote}

  Suppose the reporter's differential responsive dispositions to call
  things red are matched by those of a parrot trained to utter the
  same noises under the same stimulation. What practical capacities of
  the human distinguish the reporter from the parrot? What, besides
  the exercise of regular differential responsive dispositions, must
  one be able to \emph{do}, in order to count as having or grasping
  \emph{concepts}? ... To grasp or understand a concept is, according
  to Sellars, to have practical mastery over the inferences it is
  involved in... The parrot does not treat ``That's red'' as
  incompatible with ``That's green''\footnote{\cite{brandom2}
    pp.88-89, our emphasis.}.
\end{quote}

\NI If this claim is also accepted, then material incompatibility is
not just conceptually \emph{independent} of logical negation, but
conceptually \emph{prior}.  

\subsection{Negation as the minimal incompatible}

In \cite{brandom2} and \cite{brandom}, Brandom describes 
logical negation as a limiting form of material incompatibility:
\begin{quote}
Incompatible sentences are Aristotelian \emph{contraries}. A sentence
and its negation are \emph{contradictories}. What is the relation
between these? Well, the contradictory is a contrary: any sentence is
incompatible with its negation. What distinguishes the contradictory
of a sentence from all the rest of its contraries? The contradictory
is the \emph{minimal} contrary: the one that is entailed by all the
rest. Thus every contrary of ``Plane figure $f$ is a circle'' - for
instance ``$f$ is a triangle'', ``$f$ is an octagon'', and so on -
entails ``$f$ is \emph{not} a circle''.
\end{quote}

\NI If someone asserts that it is not the case that Pierre is the (only) King of France,
we have said very little.  There are so many different ways in which
it could be true:
\begin{itemize}
\item
The King of France might be Jacques
\item
The King of France might be Louis
\item
...
\item
There may be no King of France at all
\item
There may be no country denoted by the word ``France''
\end{itemize}
Each of these concrete propositions is incompatible with Pierre being the King of France.
To say ``It is not the case that the King of France is Pierre'' is just to claim that one of these indefinitely many concrete possibilities is true.
Negation is just the logically weakest form of incompatibility.

In the rest of this paper, we assume - without further argument - that material incompatibility is conceptually prior to logical negation.
We develop a simple
 modal logic to articulate Brandom's intuition: a language, without negation, in which we can nevertheless make incompatible claims.

\subsection{Inferences between atomic sentences}
\label{intrasentential}
So far, we have justified the claim that incompatibility is a
fundamental logical concept by arguing that incompatibility is
conceptually prior to negation.  Now incompatibility is an inferential
relation between \emph{atomic sentences}.  In this subsection, we
shall describe \emph{other} inferential relations between atomic
sentences - inferential relations that \fol{} cannot articulate (or
can only do so awkwardly), but that \cathoristic{} handles naturally.

The \emph{atomic sentences} of a natural language can be
characterised as the sentences which do not contain any other
sentences as constituent parts\footnote{Compare Russell \cite{russell}
  p.117: ``A sentence is of atomic form when it contains no logical
  words and no subordinate sentence''. We use a broader notion of
  atomicity by focusing solely on whether or not it contains a
  subordinate sentence, allowing logical words such as ``and'' \emph{as long
  as they are conjoining noun-phrases} and not sentences.}.  According
to this criterion, the following are atomic:

\begin{itemize}

\item Jack is alive
\item Jack loves Jill
\end{itemize}

\NI The following is not atomic:

\begin{quote}
  Jack is alive and Jill is dead
\end{quote}

\NI because it contains the complete sentence ``Jack is alive'' as a
syntactic constituent.  Note that, according to this criterion, the
following \emph{is} atomic, despite using ``and'':

\begin{quote}
  Jack loves Jill and Joan
\end{quote}

\NI Here, ``Jack loves Jill'' is not a syntactic constituent\footnote{To see that ``Jack loves Jill'' is not a constituent of ``Jack loves Jill and Joan'', observe that ``and'' conjoins constituents of the \emph{same syntactic type}. But ``Jack loves Jill'' is a sentence, while ``Joan'' is a noun. Hence the correct parsing is ``Jack (loves (Jill and Joan))'', rather than ``(Jack loves Jill) and Joan''.}.

There are many types of inferential relations between atomic
sentences of a natural language.  For example:

\begin{itemize}

\item ``Jack is alive'' is incompatible with ``Jack is dead''
\item ``Jack loves Jill'' implies ``Jack loves''
\item ``Jack walks slowly'' implies ``Jack walks''
\item ``Jack loves Jill and Joan'' implies ``Jack loves Jill''
\item ``Jack is wet and cold'' implies ``Jack is cold''

\end{itemize}

\NI The first of these examples involves an incompatibility relation,
while the others involve entailment relations.  A key question this
paper seeks to answer is: what is the simplest logic that can capture
these inferential relations between atomic sentences?

\subsection{Wittgenstein's vision of a logic of elementary propositions}

\NI In the \emph{Tractatus} \cite{wittgenstein-tractatus}, Wittgenstein
claims that the world is a set of atomic sentences in an idealised
logical language.  Each atomic sentence was supposed to be
\emph{logically independent} of every other, so that they could be
combined together in every possible permutation, without worrying
about their mutual compatibility.
But already there were doubts and problem cases.  He was aware that
certain  statements seemed atomic, but did not seem logically
independent:

\begin{quote}
  For two colours, e.g., to be at one place in the visual field is
  impossible, and indeed logically impossible, for it is excluded by
  the logical structure of colour. (6.3751)
\end{quote}

\NI At the time of writing the \emph{Tractatus}, he hoped that further
analysis would reveal that these statements were not really atomic.

Later, in the \emph{Philosophical Remarks} \cite{wittgenstein-remarks}, he
renounced the thesis of the logical independence of atomic
propositions.  In \S 76, talking about incompatible colour predicates,
he writes:

\begin{quote}
  That makes it look as if \emph{a construction might be possible
    within the elementary proposition}. That is to say, as if there
  were a construction in logic which didn't work by means of truth
  functions.  What's more, it also seems that these constructions have
  an effect on one proposition's following logically from another.
  For, if different degrees exclude one another it follows from the
  presence of one that the other is not present.  In that case,
  \emph{two elementary propositions can contradict one another}.
\end{quote}

\NI Here, he is clearly imagining a logical language in which there
are incompatibilities between atomic propositions. In \S 82:

\begin{quote}
  This is how it is, what I said in the Tractatus doesn't exhaust the
  grammatical rules for 'and', 'not', 'or', etc; \emph{there are rules
    for the truth functions which also deal with the elementary part
    of the proposition}.  The fact that one measurement is right
  \emph{automatically} excludes all others.
\end{quote}

\NI Wittgenstein does not, unfortunately, show us what this
language would look like.  
In this paper, we present \cathoristic{} as one way of formalising inferences
between atomic sentences.

